Applications of Technology and Large Data in Statistics Education and Statistical Graphics

### 2015-05-30 DefenseSlides/DefenseSlidesCommitteeHour.Rnw
Lots of little edits made as I practiced slides

\setbeamercovered{transparent}
\renewcommand{\ni}{\noindent}

\newcommand{\distas}[1]{\mathbin{\overset{#1}{\sim}}}%
\newcommand{\Cov}{\text{Cov}}%

@@

Primary goals:
  \begin{itemize}
    \item Introduce spatial and frequency loss
    \item Investigate binning properties
  \end{itemize}

@@

\begin{frame}
    \frametitle{Bivariate Data Visualization }
Example Data from Baseball   
  \begin{itemize}
    \item Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/})
    \item Variables are Games Played and Strike Outs
    \item Observations: 42583 MLB pitcher seasons
    \item Data from 1871-2009
  \end{itemize}
Goal: Visualize relationship between games played and strike outs

@@

Random Rectangular Binning
  \begin{itemize}
    \item Define set of bin centers: $\{x_j^\ast \text{ }|\text{ } x_j^\ast > x_{j-1}^\ast \}$ for $j \in \{1,\dots,J\}$
    \item $x_1^\ast \le x_{(1)} \le x_{(n)} \le x_J^\ast$
    \item Assign points at random to surrounding bin centers
    \item Probabilities inversely proportional do distance from bin centers
  \end{itemize}

@@

\begin{frame}
    \frametitle{Simulation Study}

@@

Data then rounded to nearest even integer to induce a \textit{data resolution} of 2 units

@@

qplot(x,y,geom="point",data=subset(alldat, record=="Fine"),facets=.~type, size=I(1)) + 
  theme_bw() + theme(aspect.ratio=1) +
  #ggtitle("Scatterplots of Simulated Data") +
  xlab("X") + ylab("Y") + 
  scale_x_continuous(breaks=c(0,25,50,75,100)) + 
  scale_y_continuous(breaks=c(0,25,50,75,100))
@

@@

Specifying a binning origin to lower spatial loss
  \begin{itemize}
    \item Avoid heavy overhangs outside data boundaries
    \item Try to align bin centers with possible data values 
  \end{itemize}

A reliable default based on the data resolution
  \begin{itemize}
    \item Recorded to nearest $\alpha_x$ units in the X dimension 
    \item Recorded to nearest $\alpha_y$ units in the Y dimension
    \item $x_{(1)}$ and $y_{(1)}$ are the minimum data values in each dimension
    \item Set binning origin at $(x_{(1)}, y_{(1)}) - (\alpha_x/2,\alpha_y/2) $
  \end{itemize}

@@

@@

How do we discretize the frequency shading?
  \begin{itemize}
    \item Add second stage of binning
    \item Use one dimensional binning algorithms on bin counts
    \item Shade ordinally based on binned frequency groups
    \item May transform counts before frequency binning \\ (i.e. log counts)
  \end{itemize}

@@

Standard Frequency Binning
  \begin{itemize}
    \item Create k frequency groups using equal bin width
  \end{itemize}  
\pause

@@

Quantile Frequency Binning
  \begin{itemize}
    \item Create k frequency groups based on empirical quantiles of raw bin counts
    \item Roughly equal number of tiles in each shade
  \end{itemize}  
\pause

@@

Frequency Loss
  \begin{itemize}
    \item bin frequencies, $c_\ell$, for spatial bins $\ell \in \{1, \dots, \mathscr{L} \}$\\
    \item binned frequencies, $b_C(c_\ell)$
  \end{itemize}  
%
\begin{eqnarray}\label{totfreqloss}
L^F = \sum_{\ell=1}^{\mathscr{L}} (c_\ell-b_c(c_\ell))^2
\end{eqnarray}  
\pause

Log Frequency Loss
  \begin{itemize}
    \item If bin frequencies, $c_\ell$, are log transformed 
    \item binned log frequencies, $b_C(\log(c_\ell))$
  \end{itemize} 
  %
\begin{eqnarray}\label{logfrequency}
L^{\log F} = \sum_{\ell=1}^{\mathscr{L}} (\log(c_\ell) - b_c(\log(c_\ell)))^2
\end{eqnarray}

@@

\center How many different shades can we distinguish simultaneously? \\ 
Research suggests 7 is max for the typical person $\Rightarrow$ use k $\le 7$

@@

Frequency Binning
  \begin{itemize}
    \item Use 4 to 7 frequency bins
    \item Use standard log freq or quantile freq binning for desired interpretation
  \end{itemize}

@@

Cronbach's $\alpha$ measures internal consistency within an item set
%
\begin{equation} \label{eq:alpha}
\alpha = K/(K-1) \cdot \left(1 - \left . \sum_{i=1}^K \V{Y_i} \right /  \V{\sum_{j=1}^K Y_j}\right),
\end{equation}

>>>>

\setbeamercovered{transparent}
\renewcommand{\ni}{\noindent}
\newcommand{\V}[1]{\text{Var}\left(#1\right)}
\newcommand{\distas}[1]{\mathbin{\overset{#1}{\sim}}}%
\newcommand{\Cov}{\text{Cov}}%

@@

Example Data from Baseball   
  \begin{itemize}
    \item Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/})
    \item Variables are Games Played and Strike Outs
    \item Observations: 42583 MLB pitcher seasons
    \item Data from 1871-2009
    \item Goal: Visualize relationship between games played and strike outs
  \end{itemize}

@@

% \begin{frame}
%     \frametitle{Bivariate Data Visualization }
% Example Data from Baseball   
%   \begin{itemize}
%     \item Sean Lahman's Baseball database (\url{http://www.seanlahman.com/baseball-archive/})
%     \item Variables are Games Played and Strike Outs
%     \item Observations: 42583 MLB pitcher seasons
%     \item Data from 1871-2009
%   \end{itemize}
% Goal: Visualize relationship between games played and strike outs
% 
% \end{frame}

@@

Random Rectangular Binning
  \begin{itemize}
    \item Define set of bin centers: $\{x_j^\ast \text{ }|\text{ } x_j^\ast > x_{j-1}^\ast \}$ for $j \in \{1,\dots,J\}$
    \item Assign points at random to surrounding bin centers
    \item Probabilities inversely proportional do distance from bin centers
  \end{itemize}

@@

\begin{frame}
    \frametitle{Simulated Data}

@@

Data pairs then rounded to nearest even integer to induce a 2X2 unit \textit{data resolution}

@@

# qplot(x,y,geom="point",data=subset(alldat, record=="Fine"),facets=.~type, size=I(1)) + 
#   theme_bw() + theme(aspect.ratio=1) +
#   #ggtitle("Scatterplots of Simulated Data") +
#   xlab("X") + ylab("Y") + 
#   scale_x_continuous(breaks=c(0,25,50,75,100)) + 
#   scale_y_continuous(breaks=c(0,25,50,75,100))
@

@@

\begin{frame}
    \frametitle{Selecting a Binning Origin}
  \center Avoid binning origins with large overhangs outside data boundaries

@@

A natural binning origin is ($x_{(1)}$, $y_{(1)}$)
  \begin{itemize}
    \item where $x_{(1)}$ and $y_{(1)}$ are the minimum data values in each dimension
    \item If data resolution is coarse, points not well centered within bins
  \end{itemize}

A better default based on the data resolution
  \begin{itemize}
    \item Recorded to nearest $\alpha_x$ units in the X dimension 
    \item Recorded to nearest $\alpha_y$ units in the Y dimension
    \item Set binning origin at $(x_{(1)}, y_{(1)}) - (\alpha_x/2,\alpha_y/2) $
  \end{itemize}

@@

How do we discretize the frequency shading?
  \begin{itemize}
    \item Create data triples $(x^\ast_\ell, y^\ast_\ell, c_\ell)$ for bins $\ell \in \{1, \dots, \mathscr{L} \}$
    \item Apply a univariate binning function to raw frequencies, $c_\ell$
    \item Shade ordinally based on binned frequencies, $b_C(c_\ell)$
    \item May transform counts before frequency binning \\ (i.e. log counts)
  \end{itemize}

@@

Standard Frequency Binning
  \begin{itemize}
    \item Create k frequency groups using equal frequency bin widths
  \end{itemize}  
\pause

@@

Quantile Frequency Binning
  \begin{itemize}
    \item Create k frequency groups with empirical quantiles of raw frequencies
    \item Roughly equal number of tiles in each shade
  \end{itemize}  
\pause

@@

Frequency Loss
  \begin{itemize}
    \item Raw frequencies, $c_\ell$, for spatial bins $\ell \in \{1, \dots, \mathscr{L} \}$\\
    \item Binned frequencies, $b_C(c_\ell)$
  \end{itemize}  
%
\begin{eqnarray}\label{totfreqloss}
L^F = \sum_{\ell=1}^{\mathscr{L}} (c_\ell-b_c(c_\ell))^2
\end{eqnarray}  
\pause

Log Frequency Loss
  \begin{itemize}
    \item If bin frequencies, $c_\ell$, are log transformed 
    \item Binned log frequencies, $b_C(\log(c_\ell))$
  \end{itemize} 
  %
\begin{eqnarray}\label{logfrequency}
L^{\log F} = \sum_{\ell=1}^{\mathscr{L}} (\log(c_\ell) - b_c(\log(c_\ell)))^2
\end{eqnarray}

@@

\center How many different shades can we distinguish simultaneously? \\ 
Typical human perception $\Rightarrow$ k $\le 7$ (Healy and Enns, 1999)

@@

Frequency Binning
  \begin{itemize}
    \item May be used to make frequency loss explicit
    \item Use 4 to 7 frequency bins
    \item Quantile frequency binning provides alternative display for skewed counts
  \end{itemize}

@@

Cronbach's $\alpha$ measures internal consistency within an item set
%
\begin{equation} \label{eq:alpha}
\alpha = K/(K-1) \cdot \left(1 - \sum_{i=1}^K \V{Y_i}  /  \V{\sum_{j=1}^K Y_j}\right)
\end{equation}
\begin{itemize}
  \item $Y_j$ denote the items $j \in \{1,\dots,K\}$ in the set
  \item Note: K=4 for sets in this survey
\end{itemize}
